The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-06 conf_article/neon.tex
Expanded the experiments

	In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan k-norms to introduce a \deleted[id=IK, comment={Redundant}]{new} family of Muon-like algorithms we name {\it Fanions}, which happen to be similar to Dion. \replaced[id=IK, comment={This rewrite clarifies the sentence structure, making it clear that the norms (not the algorithms) are being combined.}]{Subsequently, we construct a second family of algorithms, dubbed {\it F-Fanions}, based on the duals of convex combinations of Ky Fan k-norms and the Frobenius norm. One prominent member of this family is {\it F-Muason}.}{Then, we consider their convex combinations with the Frobenius norm and corresponding duals to develop a new family of algorithms we name {\it F-Fanions}, one of them being {\it F-Muon}.} We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that properly-tuned F-Muon is almost on par with Muon, \replaced[id=IK, comment={Alex, clarify this please. It's a very non-obvious and non-clear conclusion, especially when knowing that Muon itself only provides an approximate answer to the LMO. I provided an alternative phrase.}]{which questions the necessity of finding the exact solution to the problem in linear optimization oracle framework.}{which shows the flexibility in the choice of the norm and opens possibilities for easier-to-compute variants.}

@@

We introduce the family of {\it Fanions}, which consists of {\it $k$-Fanions}, lmo-based algorithms under the $\normkfk{\mM^t}^\dagger$ norms. By this terminology and the lemma, Muon is a $\min\{n,m\}$-Fanion, while Neon is an 1-Fanion. Moreover, the centralized version of rank-$r$ Dion \citep{ahn2025dioncommunicationefficientoptimizerlarge} without the error feedback and without scaling of the update, from the perspective of lmo, is actually a $r$-Fanion.

@@

\begin{table*}[t]
	\caption{lmo optimizers in Schatten $S_p$ norms and in $l_p$ norms. $g$ is the gradient. When it is a matrix, $g = \mU \mSigma \mV^\top$}
	\label{tbl:mat_vs_vec_lmo}
	\bgroup
	\def\arraystretch{1.2}
	\resizebox{\textwidth}{!}{
		\begin{tabular}{|l|c|c|c|}
			\hline
			Method                             & lmo constraint set $\mathcal D$ & lmo                                                         & Reference                            \\
			\hline
			\hline
			Normalized SGD                     & $l_2$-ball, $S_2$-ball          & $-\eta \tfrac{g}{\norm{g}_2} = -\eta \tfrac{g}{\norm{g}_F}$ & \citep{hazan2015beyond}              \\
			Momentum Normalized SGD            & Ball in $l_2$, or Ball in $S_2$ & $-\eta \tfrac{g}{\norm{g}_2} = -\eta\tfrac{g}{\norm{g}_F}$  & \citep{cutkosky2020momentum}         \\
			\hline
			SignSGD                            & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 1]{bernstein2018signsgd} \\
			Signum                             & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 3]{bernstein2018signsgd} \\
			\hdashline
			Muon                               & Ball in Spectral $S_\infty$     & $-\eta UV^\top$                                             & \citep{jordan2024muon}               \\
			\hline
			Gauss-Southwell Coordinate Descent & Ball in $l_1$                   & $-\eta \{i: g_i \geq g_k \forall k\}$                       & \citep[p. 19]{shi2016primer}         \\
			\hdashline
			Neon                               & Ball in Nuclear $S_1$           & $-\eta u_1 v_1^\top$                                        & This work                            \\
			\hline
		\end{tabular}
	}
	\egroup
\end{table*}

@@

The practical drawback of the TRLAn is the absence of its implementation in PyTorch.
\begin{table}[h!]
	\centering
	\begin{tabular}{c|c|c|c}
		\hline
		Method                              & rtol & k   & time,s \\
		\hline
		Power Iterations                    & 0.01 & 1   & 7.7    \\
		SVDS (thick-restart Lanczos method) & 0.01 & 1   & 0.18   \\
		PCA Low Rank (RSVD)                 & 0.01 & 1   & 1.15   \\
		SVDS (thick-restart Lanczos method) & 0.01 & 10  & 0.47   \\
		PCA Low Rank (RSVD)                 & 0.01 & 10  & 19.4   \\
		SVDS (thick-restart Lanczos method) & 0.01 & 100 & 1.96   \\
		PCA Low Rank (RSVD)                 & 0.01 & 100 & 170    \\
	\end{tabular}
	\caption{k-rank updated comparison}
	Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation.
	\label{tbl:matrix_methods}
\end{table}
\section{Trust Region Bounds for L-smooth functions}
First, we analyze the problem in the unstochastic case. From Corollary 1 of \citet{kovalev2025understanding}, we directly get the following result that matches lower bounds, as was noted in that article.

\begin{lemma}\label{lemma:no_noise_tr}
	To reach the precision $\min_{k=1\ldots K} \norm{\nabla f(\mX_k)}^\dagger \leq \epsilon$ by the iterations~\eqref{eq:our_update} under the conditions of \Cref{eq:L}, it is sufficient to choose the stepsize $\eta$ and the number of iterations $K$ as follows:
	\begin{equation}\label{eq:unstoch_tr}
		\eta = \cO\left(\frac{\epsilon}{L}\right),\qquad K = \cO\left(\frac{L\Delta_0}{\epsilon^2}\right).
		% \eta = \frac{\epsilon}{3L},\qquad K = \ceil*{\frac{6L\Delta_0}{\epsilon^2}}.
	\end{equation}
\end{lemma}

In the stochastic case, from Corollary of 2 of \citet{kovalev2025understanding}, we directly get the following result that once more matches lower bounds:

\begin{lemma}\label{lemma:stoch_tr}
	To reach the precision $\E{\min_{k=1\ldots K} \norm{\nabla f(\mX_k)}^\dagger} \leq \epsilon$ by ~\eqref{eq:our_update} under the assumptions \Cref{eq:variance}, \Cref{eq:L}, \Cref{eq:norm}, it is sufficient to choose the parameters as follows:
	\begin{align}
		\eta & = \cO\left(\min\left\{\frac{\epsilon}{L}, \frac{\epsilon^3}{\rho^2\sigma^2L}\right\}\right),
		\qquad
		\alpha = \cO\left(\min\left\{1, \frac{\epsilon^2}{\rho^2\sigma^2}\right\}\right),
		\\
		\label{eq:str_K_nonconvex}
		K    & = \cO\left(\max\left\{
		\frac{\rho\sigma}{\epsilon},
		\frac{\rho^3\sigma^3}{\epsilon^3},
		\frac{L\Delta_0}{\epsilon^2},
		\frac{L\Delta_0\rho^2\sigma^2}{\epsilon^4}
		\right\}\right).
	\end{align}
\end{lemma}

As the norms $\normf{\cdot}$, $\normn{\cdot}$, $\norms{\cdot}$ are almost proportional to each other when $m, n \rightarrow \infty$ (with high probability for random matrices), the expected convergence guarantees in terms of $\normf{\cdot}$ are the same (it can be easily shown by noting that $\norm{\mX}\sim\alpha\normf{\mX}$, $\norm{\nabla f(\mX)}^\dagger \sim \frac{1}{\alpha} \normf{\nabla f(\mX)}$, and expressing $L$-constant via $L_F$-constant for the Frobenius norm).

From the theory of random martices and the Marchenko-Pastur law, we get that random $\mM \in \Rmn: \mM \sim \cN(0, \sigma^2 \Rmn)$ has the following asymptotics of its norms:

Nuclear: $\sigma n \sqrt{m}$

Frobenius: $\sigma \sqrt{m n}$

Spectral: $\sigma(\sqrt{m} + \sqrt{n})$

This means that for square random matrices $n \times n$  the following asymptotics take place: $\normf{\cdot} \sim \frac{\sqrt{n}}{2}\norms{\cdot}$ and $\normn{\cdot} \sim \frac{n}{2}\norms{\cdot}$.

\section{Experiments}\label{sec:experiments}
\subsection{Randomized Linear Least Squares}
Since the provided by other authors \citep{kovalev2025understanding,riabinin2025gluon} theoretical guarantees are almost norm-independent, we have to test them in practice.

To test the bounds from \citet{kovalev2025understanding}, we construct the following L-smooth problem:

\begin{equation}\label{eq:lls}
	F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN>
\end{equation}

where $\mX \in \Rmn$, $m=10$, $n=10$, $S \in \Rmn$, $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. Spectra of $\mM$ and $\mN$ are uniformly distributed in (0, 1) interval.

It it easy to derive the gradient
\begin{equation}\label{eq:grad_lls}
	\nabla F(\mX) = \mM (\mX - \mS) \mN,
\end{equation}

Let us define $\gamma$ as $\norm{\cdot} \sim \gamma \normf{\cdot}$, which is asymptotics from the previous section. Then $\norm{\cdot}^\dagger \sim \frac{1}{\gamma}\normf{\cdot}$, as $\normf{\cdot}^\dagger = \normf{\cdot}$. Hence, $\norm{\cdot} \sim \gamma^2 \norm{\cdot}^\dagger$

Then $\norm{\nabla f(\mX) - \nabla f(\mY)}^\dagger = \norm{\mM (\mX - \mY) \mN}^\dagger \leq \norm{\mM}^\dagger \norm{\mN}^\dagger \norm{\mX - \mY}^\dagger \sim \norm{\mM}^\dagger \norm{\mN}^\dagger \gamma^2 \norm{\mX - \mY}$, and $L \sim \gamma^2 \norm{\mM}^\dagger \norm{\mN}^\dagger$.

Now for the known norms:

Frobenius (NSGD): $L = \normf{M} \normf{N}$

Spectral (Muon): $\gamma \sim \frac{2}{\sqrt{n}} \implies L \sim \frac{2}{n}\normn{M}\normn{N}$

Nuclear (Neon): $\gamma \sim \sqrt{n} \implies L \sim n\norms{M}\norms{N}$

We take learning rate $\eta$ and iteration number $K$ from \cref{eq:unstoch_tr}.

We are interested to get $\epsilon_2=$ 1e0 precision in $\norms{\cdot}$. Hence, we target $\epsilon = \epsilon_2$ for the nuclear norm (Neon), $\epsilon = \frac{n}{2}$ for the spectral norm (Muon), and $\epsilon = \frac{\sqrt{n}}{2}$ for the Frobenius norm (NSGD).
With that, we get the following lrs: Muon lr: 2.33E-04, Neon lr: 2.10e-02, Frobenius (NSGD) lr: 1.75e-02, and the following number of iterations: Muon iter: 68, Neon iter: 11919, Frobenius iter: 323. L-constants are 171.4, 47.7, and 16.2 respectively. As we will see, these results do not relate well to the practice.

To make the problem more real-world, we do not use known information about smoothness. We run NormalizedSGD, Muon, F-Muon with $\alpha=1/2$, Neon, and F-Neon with $\alpha=1/2$ for 5 000 iterations with learning rate = 1e-2, which has been close to computed theoretically above. The results are in \cref{fig:lls}.

Though Neon displays the worst convergence in terms of loss, it is a leader in terms of the norm of the gradient. This might signify that the known bounds are not sufficient even in the case of a convex optimization: lmo-based algorithms may converge well by these bounds, but be impractical like Neon.

This experiment may also show that F-Muon may be more robust under general conditions than Muon.

\begin{figure}[h!]
	\centering
	\begin{subfigure}[t]{0.48\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/simple_lls/loss_vs_iteration_50x50.pdf}
		\caption{Loss}
		\label{fig:lls_loss}
	\end{subfigure}
	\hfill
	\begin{subfigure}[t]{0.48\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/simple_lls/gradient_spectral_norm_vs_iteration_50x50.pdf}
		\caption{Spectral norm of the gradient}
		\label{fig:lls_grad_norm}
	\end{subfigure}
	\caption{Linear least squares problem for a 50 x 50 matrix}
	\label{fig:lls}

\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{figs/fstardual_cifar.pdf} % adjust width as needed
	\caption{Visualization of lmo balls for Muon and F-Muon.}
	\label{fig:cifar_ball}
\end{figure}

@@

\subsection{Modded NanoGPT speedrun}
We test F-Muon on NanoGPT speedrun \citep{modded_nanogpt_2024}. For $\alpha = 0.5$, the optimal parameters are {\tt lr=0.07, momentum=0.95}, while for Muon they were  {\tt lr=0.05, momentum=0.95}. After testing for 1750 steps, as it has been done on the speedruns, we get 3.281 cross-entropy loss, while on Muon, it falls below the target threshold 3.28 reaching 3.279.

@@

\section{Formal assumptions}
Here we restate assumptions from \citet{kovalev2025understanding} that are used in the theorems from the main part of the article.

	{\bf Lipschitz continuous gradient.}
We assume that the gradient $\nabla f(\cdot)$ is Lipschitz continuous with respect to the norm $\norm{\cdot}$, that is, the following inequality holds:
\begin{aequation}\label{eq:L}
	\norm{\nabla f(\mX) - \nabla f(\mX')}^\dagger \leq L \norm{\mX-\mX'}
	\quad\text{for all}\;
	\mX, \mX' \in \Rmn,
\end{aequation}
where $L > 0$ is the gradient Lipschitz constant.

	{\bf Stochastic gradient estimator.}
We assume access to a stochastic estimator $g(\cdot\,;\xi)\colon \Rmn \to \Rmn$ of the gradient $\nabla f(\cdot)$, where $\xi \sim \cD$ is a random variable sampled from a probability distribution $\cD$. We assume that the stochastic gradient estimator $g(\cdot;\xi)$ is unbiased and has bounded variance, that is, the following relations hold:
\begin{aequation}\label{eq:variance}
	\Ed{\xi \sim \cD}{g(\mX;\xi)} = \nabla f(\mX)
	\quad\text{and}\quad
	\Ed{\xi \sim \cD}{\sqnf{g(\mX;\xi) - \nabla f(\mX)}} \leq \sigma^2
	\quad\text{for all}\;
	\mX \in \Rmn,
\end{aequation}
where $\sigma > 0$ is a positive variance parameter, and $\normf{\cdot}$ is the standard Euclidean, i.e. Frobenius, norm induced by the inner product $\<\cdot,\cdot>$, i.e., $\normf{\mX} = \sqrt{\<\mX,\mX>}=\sqrt{\tr(\mX^\top \mX)}$.

It is important to highlight that while \Cref{eq:L} uses the dual norm $\norm{\cdot}^\dagger$ to measure the difference between the gradients, the variance in \Cref{eq:variance} is measured with respect to the Frobenius norm $\sqnf{\cdot}$, which is necessary to properly utilize the unbiasedness property of the stochastic gradient estimator $g(\cdot;\xi)$. Therefore, we need to provide a connection between these norms using the following inequality:
\begin{aequation}\label{eq:norm}
	\norm{\mX}^\dagger \leq \rho\cdot\normf{\mX}
	\quad\text{for all}\;
	\mX \in \Rmn,
\end{aequation}
where $\rho > 0$ is a positive constant.

@@

\begin{figure}[h!]
	\centering
	\begin{subfigure}[b]{0.48\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/fstardualball.pdf}
		\caption{lmo balls for F-Muon for different $\alpha$}
		\label{fig:fstardual}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.48\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/ftwodualball.pdf}
		\caption{lmo ball for F-Neon for different $\alpha$}
		\label{fig:ftwodual}
	\end{subfigure}
	\caption{Balls in the duals to F* and F2 norms for different $\alpha$}
	\label{fig:fduals}
\end{figure}

@@

\begin{figure}[h!]
	\centering
	\begin{subfigure}[b]{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/KyFan.pdf}
		\caption{Ky Fan 2-norm}
		\label{fig:kyfan}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/KyFanDual.pdf}
		\caption{Dual to Ky Fan 2-norm}
		\label{fig:kyfandual}
	\end{subfigure}
	\caption{Ky Fan 2-norm and its dual}
	\label{fig:kyfan_combined}
\end{figure}

@@

@@

\iffalse
	\section{Idea}
	\iffalse
		The goal of the project is to make variations on Muon to speed it up. Recently, authors of \cite{bernstein2024oldoptimizernewnorm} have proposed to derive update step for an optimizer as the solution of the certain optimization problem. This approach can be utilized to derive Muon \cite{jordan2024muon}, a novel algorithm for fast training of neural networks. Instead of using spectral norm, used to derive Muon, we utilize nuclear norm to produce a new optimization algorithm.
	\fi
	\subsection{Problem (Project description)}

	In this subsection, we provide a more detailed description of our idea and formulate it as a mathematical problem. The authors of \cite{bernstein2024oldoptimizernewnorm} suggest obtaining the update step as a solution to the optimization problem:
	\begin{equation}
		\langle g, \delta w \rangle + \lambda \norm{\delta w}^2 \to \min_{\delta w}\,,
	\end{equation}
	where $w$ is the weight vector, $g$ is a gradient-like vector (e.g., obtained via momentum SGD), and $\norm{\cdot}$ represents a certain norm. Many popular optimizers, such as Adam (with exponential moving average disabled) and vanilla SGD, can be cast within this framework \cite{bernstein2024oldoptimizernewnorm}.

	In large language models, most weights are structured as matrices, which offers additional opportunities for optimization. Let $W$ be the weight matrix of a linear layer, and $G$ be a gradient-like matrix. Then, the update step $\delta W$ can be obtained as a solution to the optimization problem:
	\begin{equation}\label{eqn:opt_problem_mat}
		\langle G, \delta W \rangle + \lambda \norm{\delta W}^2 \to \min_{\delta W}\,,
	\end{equation}
	where $\norm{\cdot}$ denotes a certain matrix norm. By setting this norm to the RMS-to-RMS norm (a scaled version of the spectral norm), we recover the Muon optimizer \cite{bernstein2025deriving, bernstein2024oldoptimizernewnorm} with an update step defined by:
	\begin{equation}\label{eqn:update_muon}
		\delta W = - \frac{1}{\lambda}\sqrt{\frac{n}{m}}UV^T\,,
	\end{equation}
	where $m$ is the input dimension of the layer, $n$ is the output dimension, and $U$ and $V$ are obtained from the singular value decomposition of the gradient matrix $G = U \Sigma V$.

	Motivated by the recent achievements of the Muon optimizer (e.g., \cite{liu2025muon}), we consider alternative choices of norms, specifically the nuclear norm $\norm{\cdot}_*$ and a custom $F*$ norm, given by
	\begin{equation}\label{eqn:F_star}
		\norm{X}_{F*}^2 = \frac{\norm{X}_F^2 + \norm{X}_*^2}{2}\,,
	\end{equation}
	where $\norm{\cdot}_F$ denotes the Frobenius norm.

	Using the nuclear norm in \eqref{eqn:opt_problem_mat} leads to a rank-one update of the weight matrices:
	\begin{equation}\label{eqn:update_star}
		\delta W = -\frac{1}{2\lambda} u_1 \sigma_1 v_1^T\,,
	\end{equation}
	where $\sigma_1$ is the largest singular value, and $u_1$ and $v_1$ are the corresponding singular vectors. We expect one iteration of this method to be significantly faster than one iteration of Muon.

	Another choice is the $F*$ norm. With this choice, \eqref{eqn:opt_problem_mat} yields
	\begin{equation}\label{eqn:update_F_star}
		\delta W = -\frac{1}{\lambda}UDV^T
	\end{equation}
	with $D = \text{diag}(d_i)$, where $d_i = [\sigma_i - \tau]_+$, and $\tau$ is given by
	\begin{equation}
		\sum_{i} [\sigma_i - \tau]_+ = \tau\,.
	\end{equation}
	We anticipate that the method with this update step will perform well with large batch sizes.

	In this article we show how one can quickly compute weight updates defined by \eqref{eqn:update_star} or \eqref{eqn:update_F_star}. Then we finilize the method by adding momentum and test their performance against those of Muon at training multilayer perceptron and transformer. The results will be fast algorithm, which we will convert into a new optimizer classes for PyTorch, as was done with Muon.

	% Removed template instructions sections
	\section{Related Work}
	Our review primarily focuses on the Muon and Shampoo optimizers, as our algorithm extends the ideas used to derive these methods. We highlight the advantages and disadvantages of these approaches, the unique effects they introduce, and compare them to Neon.

	\subsection{Muon optimizer}

	In the previous section, we described the theoretical foundation behind the weight update step in the Muon optimizer, but we did not discuss how to obtain the required matrices in practice. The update step is defined by \eqref{eqn:update_muon}, which requires $UV^T$ with $U$ and $V^T$ from the singular value decomposition of the gradient-like matrix $G$. A naive solution would involve computing the SVD of $G$ and constructing the required expression. However, the developers of the Muon optimizer introduced a workaround using Newton-Schulz iterations \cite{jordan2024muon}. The Newton-Schulz iterations from the original article \cite{jordan2024muon} require 10 matrix-matrix multiplications to achieve the desired accuracy. The asymptotic complexity of such an operation is identical to that of SVD and equals $O(mn \min\{m, n\})$ for an $m \times n$ matrix. Nevertheless, matrix multiplication on modern GPUs can be performed much more efficiently.

	The performance of Muon in training large language models was tested \cite{liu2025muon} against AdamW. The testing demonstrated excellent performance by Muon, which was approximately 2 times more efficient in terms of FLOPs required to reach a certain loss value. This is even more remarkable when considering the cost of one iteration: Muon requires an additional $O(mn \min\{m, n\})$ FLOPs per $m \times n$ matrix, while AdamW needs only $O(mn)$.

	Another interesting discovery about the Muon optimizer is that it accelerates grokking \cite{tveit2025muonoptimizeracceleratesgrokking}. In the test problem, Muon achieved grokking significantly faster than AdamW in terms of passed epochs, with a mean grokking epoch of 102.89 for Muon and 153.09 for AdamW. The authors suggest that this may be due to the fact that Muon stimulates broader exploration by orthogonalizing the gradient matrix, thus avoiding memorization.

	Recently, theoretical guarantees for Muon convergence have been derived \cite{li2025noteconvergencemuon}. In particular, in the $L$-smooth convex case, it achieves $O(1/T^{\frac{1}{2}})$ (with full gradient) and $O(1/T^{\frac{1}{4}})$ (with stochastic gradient) bounds on the Frobenius norm of the gradient or the mathematical expectation of the gradient norm, respectively, where $T$ is the number of iterations.

	\subsection{Shampoo optimizer}

	Another optimizer that exploits the matrix (and even tensor) structure of weights in neural networks is the Shampoo optimizer. We avoid a detailed description of this method here and refer the reader to the original article \cite{gupta2018shampoopreconditionedstochastictensor}, but we outline the key properties of Shampoo and its relation to the Muon optimizer.

	The Shampoo optimizer uses left and right preconditioning for the gradient-like matrix, leveling its spectrum. The preconditioners are computed from the exponentially averaged gradients, and their computation requires $O(n^3 + m^3)$ per $m \times n$ matrix. This exponential averaging is a key feature that provides several distinct interpretations for the preconditioners. They can be viewed as an approximation of the Gauss-Newton component of the Hessian or the first step of the power iteration algorithm for computing the optimal Kronecker product approximation \cite{morwani2025a}. With exponential averaging turned off, the update step of Shampoo can be simplified and becomes identical to that of Muon \cite{jordan2024muon}.

	Convergence analysis is presented in the original article \cite{gupta2018shampoopreconditionedstochastictensor}. Shampoo achieves $O(1/T^{\frac{1}{2}})$ convergence for the loss function value in the $L$-smooth convex case, where $T$ is the number of iterations.

	\subsection{Synthesis and Neon's position}

>>>>

	In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the duals to the Ky Fan k-norms to introduce a \deleted[id=IK, comment={Redundant}]{new} family of Muon-like algorithms we name {\it Fanions}, which happen to be similar to Dion. \replaced[id=IK, comment={This rewrite clarifies the sentence structure, making it clear that the norms (not the algorithms) are being combined.}]{Subsequently, we construct a second family of algorithms, dubbed {\it F-Fanions}, based on the duals of convex combinations of Ky Fan k-norms and the Frobenius norm. One prominent member of this family is {\it F-Muon}.}{Then, we consider their convex combinations with the Frobenius norm and corresponding duals to develop a new family of algorithms we name {\it F-Fanions}, one of them being {\it F-Muon}.} We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that properly-tuned F-Muon is almost on par with Muon, \replaced[id=IK, comment={Alex, clarify this please. It's a very non-obvious and non-clear conclusion, especially when knowing that Muon itself only provides an approximate answer to the LMO. I provided an alternative phrase.}]{which questions the necessity of finding the exact solution to the problem in linear optimization oracle framework.}{which shows the flexibility in the choice of the norm and opens possibilities for easier-to-compute variants.}

@@

We introduce the family of {\it Fanions}, which consists of {\it Fanion-$k$}, lmo-based algorithms under the $\normkfk{\mM^t}^\dagger$ norms. By this terminology and the lemma, Muon is a Fanion-$\min\{n,m\}$, while Neon is an Fanion-1. Moreover, the centralized version of rank-$r$ Dion \citep{ahn2025dioncommunicationefficientoptimizerlarge} without the error feedback and without scaling of the update, from the perspective of lmo, is actually a Fanion-$r$.

@@

\begin{table*}[t]
	\caption{lmo optimizers in Schatten $S_p$ norms and in $l_p$ norms. $g$ is the gradient. When it is a matrix, $g = \mU \mSigma \mV^\top$}
	\label{tbl:mat_vs_vec_lmo}
	\bgroup
	\def\arraystretch{1.2}
	\resizebox{\textwidth}{!}{
		\begin{tabular}{|l|c|c|c|}
			\hline
			Method                             & lmo constraint set $\mathcal D$ & lmo                                                         & Reference                            \\
			\hline
			\hline
			Normalized SGD                     & $l_2$-ball, $S_2$-ball          & $-\eta \tfrac{g}{\norm{g}_2} = -\eta \tfrac{g}{\norm{g}_F}$ & \citep{hazan2015beyond}              \\
			Momentum Normalized SGD            & Ball in $l_2$, or Ball in $S_2$ & $-\eta \tfrac{g}{\norm{g}_2} = -\eta\tfrac{g}{\norm{g}_F}$  & \citep{cutkosky2020momentum}         \\
			\hline
			SignSGD                            & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 1]{bernstein2018signsgd} \\
			Signum                             & Ball in Max-norm $l_\infty$     & $-\eta \sign(g)$                                            & \citep[Thm. 3]{bernstein2018signsgd} \\
			\hdashline
			Muon                               & Ball in Spectral $S_\infty$     & $-\eta UV^\top$                                             & \citep{jordan2024muon}               \\
			\hline
			Gauss-Southwell Coordinate Descent & Ball in $l_1$                   & $-\eta \sum_{i \in \argmax|g_i^t|} \sign(g_i^t)e_i$                       & \citep[p. 19]{shi2016primer}         \\
			\hdashline
			Neon                               & Ball in Nuclear $S_1$           & $-\eta u_1 v_1^\top$                                        & This work                            \\
			\hline
		\end{tabular}
	}
	\egroup
\end{table*}

@@

The practical drawback of the TRLAn is the absence of its implementation in PyTorch.
\begin{table}[h!]
	\centering
	\begin{tabular}{c|c|c|c}
		\hline
		Method                              & rtol & k   & time,s \\
		\hline
		Power Iterations                    & 0.01 & 1   & 7.7    \\
		SVDS (thick-restart Lanczos method) & 0.01 & 1   & 0.18   \\
		PCA Low Rank (RSVD)                 & 0.01 & 1   & 1.15   \\
		SVDS (thick-restart Lanczos method) & 0.01 & 10  & 0.47   \\
		PCA Low Rank (RSVD)                 & 0.01 & 10  & 19.4   \\
		SVDS (thick-restart Lanczos method) & 0.01 & 100 & 1.96   \\
		PCA Low Rank (RSVD)                 & 0.01 & 100 & 170    \\
	\end{tabular}
	\caption{k-rank updated comparison}
	Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation.
	\label{tbl:matrix_methods}
\end{table}

\section{Experiments}\label{sec:experiments}
\subsection{Randomized Linear Least Squares}
We first compare F-Fanions on the following L-smooth problem:

\begin{equation}\label{eq:lls}
	F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\
\end{equation}

where $\mX \in \Rmn$, $m=500$ and $n=500$ are typical dimensions of a neural network weight matrix, $S \in \Rmn$, $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed in the (0, 1) interval. We set $\mS=0$, and $\mX^0 \sim 0.1 \gN(0, \mI)$.

We run different Fanions and their respective F-Fanions with $\alpha=1/2$: Neon (Fanion-1), Fanion-2, Fanion-5, Fanion-100, and Muon (Fanion-500). We test them against Normalized SGD, which is also a F-Fanion with $\alpha = 0$ and an arbitrary $k$.

Since theoretical bounds \citep{kovalev2025understanding, riabinin2025gluon} rely on a very loose norm bound $\norm{\cdot}\le \rho \normf{\cdot}$, we do not derive learning rate from smoothness. Rather, we set the learning rate of $1/\sqrt{T}$ for each optimizer and the momentum of 0.6. 12 000 iterations for each algorithm ensure that the rate at some point gets low enough for each algorithm to converge. The results are presented in \cref{fig:lls} and ~\cref{sec:lls_plots_section}.

In terms of the loss minimization, F-Muon had the fastest convergence. In terms of the Frobenius norm of the gradient, the lowest was observed for NSGD. Then go F-Neon (F-Fanion-1), F-Fanion-2, F-Fanion-5, F-Fanion-100, and F-Muon (Fanion-500). It is noticeable that all F-Fanions have lower Frobenius norms of the gradients than their respective Fanioins. The same behavior will be recorded in the next section.

\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{0.8\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figs/simple_lls/loss_vs_iteration_500x500.pdf}
        \caption{The loss}
        \label{fig:lls_loss}
    \end{subfigure}
    \vspace{1em} % optional: add vertical spacing between subfigures
    \begin{subfigure}[t]{0.8\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figs/simple_lls/gradient_frobenius_norm_vs_iteration_500x500.pdf}
        \caption{The Frobenius norm of the gradient}
        \label{fig:lls_fro_grad_norm}
    \end{subfigure}
    \caption{Linear least squares problem for a 500x500 matrix}
    \label{fig:lls}
\end{figure}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{figs/norms/fstardual_cifar.pdf} % adjust width as needed
	\caption{Visualization of lmo balls for Muon and F-Muon.}
	\label{fig:cifar_ball}
\end{figure}

@@

\subsection{NanoGPT speedrun}
We test F-Muon on NanoGPT speedrun \citep{modded_nanogpt_2024}. For $\alpha = 0.5$, the optimal parameters are {\tt lr=0.07, momentum=0.95}, while for Muon they were  {\tt lr=0.05, momentum=0.95}. After testing for 1750 steps, as it has been done on the speedruns, we get 3.281 cross-entropy loss, while on Muon, it falls below the target threshold 3.28 reaching 3.279. However, this difference is negligible, if one looks at \cref{fig:gpt_loss}. It is even more striking, considering the fact that F-Muon is an average of Muon and NSGD, and the later performed quite poorly.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.7\textwidth]{figs/nanogpt/val_loss_vs_step.pdf}
	\caption{The validation loss for NanoGPT}
	\label{fig:gpt_loss}
\end{figure}

@@

@@

\begin{figure}[h!]
	\centering
	\begin{subfigure}[b]{0.48\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/norms/fstardualball.pdf}
		\caption{lmo balls for F-Muon for different $\alpha$}
		\label{fig:fstardual}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.48\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/norms/ftwodualball.pdf}
		\caption{lmo ball for F-Neon for different $\alpha$}
		\label{fig:ftwodual}
	\end{subfigure}
	\caption{Balls in the duals to F* and F2 norms for different $\alpha$}
	\label{fig:fduals}
\end{figure}

@@

\begin{figure}[h!]
	\centering
	\begin{subfigure}[b]{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/norms/KyFan.pdf}
		\caption{Ky Fan 2-norm}
		\label{fig:kyfan}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\linewidth}
		\centering
		\includegraphics[width=\linewidth]{figs/norms/KyFanDual.pdf}
		\caption{Dual to Ky Fan 2-norm}
		\label{fig:kyfandual}
	\end{subfigure}
	\caption{Ky Fan 2-norm and its dual}
	\label{fig:kyfan_combined}
\end{figure}

@@

\section{More plots for Linear Least Squares}
\label{sec:lls_plots_section}

\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{0.6\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figs/simple_lls/gradient_spectral_norm_vs_iteration_500x500.pdf}
        \caption{The spectral norm of the gradient}
        \label{fig:lls_op_grad_norm}
    \end{subfigure}
    \vspace{1em} % optional: add vertical spacing between subfigures
    \begin{subfigure}[t]{0.6\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figs/simple_lls/gradient_nuclear_norm_vs_iteration_500x500.pdf}
        \caption{The nuclear norm of the gradient}
        \label{fig:lls_nuclear_grad_norm}
    \end{subfigure}
    \vspace{1em} % optional: add vertical spacing between subfigures
    \begin{subfigure}[t]{0.6\linewidth}
        \centering
        \includegraphics[width=\linewidth]{figs/simple_lls/loss_vs_time_500x500.pdf}
        \caption{The loss over time}
        \label{fig:lls_loss_time}
    \end{subfigure}
    \caption{More images for Linear least squares problem for a 500x500 matrix}
    \label{fig:app_lls}
\end{figure}

@@
